Flexible Correlation Structures for Random Effects
========================================================

```{r}
## setwd("C:/lme4")
## load_all()
options(digits=3)
library(devtools)
library(ggplot2)
library(mvtnorm)
#load_all()

flexLmer <- function(formula, 
                     data, 
                     specials = c("cs", "d", "ar1d"), 
                     verbose = 0L){
	#split off reGenerator terms:
	frmlterms <- terms(formula, specials=specials)
	termnames <- attr(frmlterms, "variables")
	where.specials <- unlist(attr(frmlterms, "specials"))-1 #offset by one since response is counted
	lmerformula <- formula(paste(formula[[2]], "~", 
								 paste(attr(frmlterms, "term.labels")[-where.specials], collapse="+")))
	reGenerators <- as.formula(paste("~",
							paste(attr(frmlterms, "term.labels")[where.specials], collapse="+")))
	# see example(modular)
	lmod <- lFormula(lmerformula, data, reGenerators=reGenerators)
	devfun <- do.call(mkLmerDevfun, lmod)
	opt <- optimizeLmer(devfun, verbose=verbose)
	#FIXME: this should not be necessary...
	environment(devfun)$pp$theta <- opt$par
	list(model=mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr),
		 opt=opt, devfun=devfun)
}
````

### Diagonal random effects:
Uncorrelated random effects for different factor levels without having to manually specify
dummies for each level:

Example data: Different variances for different treatment levels
````{r}
## sim & plot data:
set.seed(2323)
nsubj <- 40
nreps <- 5
ntreat <- 3
n <- ntreat*nsubj*nreps
data <- expand.grid(treatment=factor(1:ntreat), 
					reps=1:5, 
					subject=factor(1:nsubj))
sd.treat <- 1:ntreat
sd.eps <- .5
beta <- (3*1:ntreat)
data <- within(data, {
	mu <- model.matrix(~treatment, data)%*%beta + 
		model.matrix(~0+subject:treatment, data) %*% rnorm(nsubj*ntreat, sd=rep(sd.treat, e=nsubj))
	y <- mu + rnorm(n, sd=sd.eps)
}) 
(p <- qplot(x=subject, y=y, col=treatment, data=data) + theme_bw() + coord_flip()) 
````
Fit & check results (estimates overlaid as crosses):
````{r}
m <- flexLmer(y ~ treatment + d(~(0+treatment|subject)), data)
mOld <- lmer(y ~ treatment + (0+treatment|subject), data)
# --> sd(treatment2)!

p + geom_point(aes(y=fitted(m$model)), shape=4, size=5)
````
True variance components: [`r c(sd.treat, sd.eps)`]

Estimated: [`r c(m$opt$par*sigma(m$model), sigma(m$model))`]

True beta: [`r beta`]

Estimated: [`r m$model@beta`]

Compare:
````{r}
mOld
````

### Variance homogeneous random effects:

Uncorrelated random effects with *same* variance for different factor levels.
````{r}
## sim & plot 
set.seed(1480)
nsubj <- 40
nreps <- 10
ntreat <- 4
n <- ntreat*nsubj*nreps
data <- expand.grid(treatment=factor(1:ntreat), 
					reps=1:nreps, 
					subject=factor(1:nsubj))
sd.treat <- 2
sd.eps <- .2
beta <- 1:ntreat

data <- within(data, {
	mu <- model.matrix(~ treatment, data)%*%beta + 
		model.matrix(~0+subject:treatment, data) %*% rnorm(ntreat*nsubj, sd=sd.treat)
	y <- mu + rnorm(n, sd=sd.eps)
}) 
(p2 <- qplot(x=subject, y=y, col=treatment, group=subject, data=data) + theme_bw()+ coord_flip())
````

Fit & check results:
````{r}
m2 <- flexLmer(y ~ treatment + d(~(0+treatment|subject), iid=TRUE), data)
m2Old <- lmer(y ~ treatment + (0+treatment|subject), data)

p2 + geom_point(aes(y=fitted(m2$model)), shape=4, size=5)
````

True variance components: [`r c(sd.treat, sd.eps)`]

Estimated: [`r c(m2$opt$par*sigma(m2$model), sigma(m2$model))`]

True beta: [`r beta`]

Estimated: [`r m2$model@beta`]

Compare:
````{r}
m2Old
````
... way too many theta's for these data.


### Compound symmetry random effects (heterogeneous variances):

I.e., Correlated random effects with *same* correlation between different effects.

````{r}
## sim & plot 
set.seed(2013)
nsubj <- 30
nreps <- 10
ndose <- 4
n <- nsubj*nreps*ndose
data <- expand.grid(reps=factor(1:nreps),
					dose=1:ndose,
					subject=factor(1:nsubj))

# set up cov.mat for random effects 
# (i.e., variance increases with dose, subject-deviations positively correlated):
sd.dose <- 1:ndose
cor <- .5
S <- {
	C <- diag(ndose)
	C[upper.tri(C)] <- C[lower.tri(C)] <- cor
	sd.dose*t(sd.dose*C)
}
sd.eps <- 1
beta <- 2

data <- within(data, {
	y <- dose*beta + 
		model.matrix(~0+as.factor(dose):subject)%*%
		as.vector(replicate(nsubj, rmvnorm(1, sigma=S)[1,]))  +
		rnorm(n, sd=sd.eps)
	doseF <- as.factor(dose)
	subjectreps <- subject:reps
})
(p3 <- qplot(x=dose, y=y, col=subject, data=data, geom=c("point", "line"), alpha=.2, facets=~reps) +
 	theme_bw() + theme(legend.position="none"))
````


Fit & check results:
````{r}
m3 <- flexLmer(y ~ dose + cs(~(0+doseF|subject)), data)
m3Old <- lmer(y ~ dose + (0+doseF|subject), data)

#true  random effect covariance:
S

#estimated: 
(estS <- with(environment(m3$devfun),{
		Lt <- pp$Lambdat
		Lt@x <- pp$thfun(m3$opt$par)
		crossprod(Lt[1:ndose, 1:ndose])
	})*sigma(m3$model)^2) 
````
True correlation: `r cor`, estimated as `r m3$opt$par[ndose+1]`.


True error sd: `r sd.eps`, estimated as `r sigma(m3$model)`.

True beta: [`r c(0, beta)`], estimated as [`r m3$model@beta`].

Compare:
````{r}
m3Old
````
sd(doseF2) is way off!

### Compound symmetry random effects (homogeneous variances):

I.e., correlated random effects with *identical* correlations between different effects.

````{r}
## sim & plot 
set.seed(2121)
nsubj <- 50
nreps <- 5
ndose <- 5
n <- nsubj*nreps*ndose
data <- expand.grid(reps=factor(1:nreps),
					dose=1:ndose,
					subject=factor(1:nsubj))

# set up cov.mat for random effects 
# (i.e., variance constant for doses, subject-deviations positively correlated):
sd.dose <- 2
cor <- .8
S <- {
	C <- diag(ndose)
	C[upper.tri(C)] <- C[lower.tri(C)] <- cor
	sd.dose*t(sd.dose*C)
}
sd.eps <- 1
beta <- 2

data <- within(data, {
	y <- dose*beta + 
		model.matrix(~0+as.factor(dose):subject)%*%
		as.vector(replicate(nsubj, rmvnorm(1, sigma=S)[1,]))  +
		rnorm(n, sd=sd.eps)
	doseF <- as.factor(dose)
	subjectreps <- subject:reps
})
(p4 <- qplot(x=dose, y=y, col=subject, data=data, geom=c("point", "line"), alpha=.2, facets=~reps) +
 	theme_bw() + theme(legend.position="none"))
````
NB: for this data with such a strong correlation between the dose-specific random intercepts
simply doing `~(1|subject)` would probably make more sense...

Fit & check results:
````{r}
m4 <- flexLmer(y ~ dose + cs(~(0+doseF|subject), het=FALSE), data)
m4Old <- lmer(y ~ dose + (0+doseF|subject), data)
````
True variance components & correlation: `r c(sd.dose, cor, sd.eps)`

Estimates: 
`r c(m4$opt$par[1]*sigma(m4$model), m4$opt$par[2], sigma(m4$model))`

True betas: `r c(0, beta)`

Estimates: `r m4$model@beta`

Compare:
````{r}
m4Old
````


### Simple AR(1)

Subjects with multiple measurements at same timepoint, errors autocorrelated over time on subject level:
````{r}
set.seed(2352)
nsubj <- 20
nreps <- 2
length <- 20
n <- nsubj*nreps*length
fr <- expand.grid(
	time=1:length,
	id=factor(1:nsubj),
	rep=factor(1:nreps))

rho <- .8
sd.b <- 3
b <- sd.b * as.vector(replicate(nsubj, arima.sim(n=length, model=list(ar=c(rho)))))				 
sd.eps <- 1
beta <- .5
fr <- within(fr,{
	mu <- beta*time
	y <- mu + model.matrix(~0+as.factor(time):id)%*%b + 
		rnorm(n, sd=sd.eps)
})	
(p5 <- qplot(y=mu, x=time, data=fr, geom="line") + facet_wrap(~id) + 
	geom_point(aes(y=y, col=rep)))
````


Fit & check results:
````{r}
m5 <-  flexLmer(y ~ time + ar1d(~(time|id)), data=fr)
````
True variance components & correlation: `r c(sd.b, rho, sd.eps)`

Estimates: 
`r c(m5$opt$par[1]*sigma(m5$model), m5$opt$par[2], sigma(m5$model))`

... scale of AR(1) seems to get overestimated all the time...?!? 

True betas: `r c(0, beta)`

Estimates: `r m5$model@beta`

... look a little weird, too. Does not seem stable yet.




### TODOs
- **for all of these cases I get the 'assertion failure' crash reported earlier if I set `nreps <- 1`!!**
- fix need for manual assignment of `theta` into `environment(devfun)` 
- `VarCorr` methods etc. for sensible output and summaries
- more structures: 
    - AR(1), at least for discrete time
    - effects with fixed correlation structure $C$: $b_i \sim N_k(0, \sigma^2 C)$, where $C$ comes from some measure of spatial/temporal/genetic/... distance.
    - diagonal covariance where $\sigma_{ij}=\exp(x_{ij}\beta)$ or similar, i.e. covariate-dependent. 

### System setup:
Using 'flexLambda' branch!
````{r}
Sys.Date()
sessionInfo()
````
