---
title: "diagnosing/understanding order dependence in `lme4` models"
date: "`r Sys.time()`"
---

This example was sent by Laura Chihara.
Construct combined data set: this uses `MathAchSchool` and `MathAchieve` from the `nlme` package, dropping some variables and renaming others in the combined data set.

```{r pkgs_utils,message=FALSE}
library(lme4)
library(GGally)
library(glmmTMB)
th <- function(obj) getME(obj,"theta")
singParams <- function(obj) {
    gg <- th(obj)
    names(gg)[abs(gg)<1e-5]
}
```

```{r make_data}
data("MathAchSchool", package="nlme")
data("MathAchieve",package="nlme")
MathScoresAll <- merge(MathAchSchool,MathAchieve,
                       by="School")
## rename
names(MathScoresAll) <- c("School","Size","Type","Academic",
                          "D","H","AveSES","Minority","Gender","SES",
                          "Math","X")
MathScoresAll$sizeScaled <- scale(MathScoresAll$Size)
## reorder
MathScoresAll <- subset(MathScoresAll,
                        select=c(School,Math,Gender,Minority,
                                 SES,Size,Type,Academic,AveSES,
                                 sizeScaled))
```

Set up formulas: the only difference between these is that they put the *terms* in the random effect in a different order.
```{r forms}
fixedpars <- c("Gender","SES","Minority","sizeScaled",
               "Type","Academic","AveSES")
ranterms <- c("Gender","Minority","SES")
tmpf <- function(rord=1:3) {
    reformulate(c(fixedpars,
                  sprintf("(%s|School)",
                          paste(ranterms[rord],collapse="+"))),
                response="Math")
}
(form.A <- tmpf())
(form.B <- tmpf(c(1,3,2)))
```

Fit both models:

```{r lme4_fit_REML,cache=TRUE}
math.lmerA <- lmer(form.A, data = MathScoresAll, REML=TRUE)
math.lmerB <- lmer(form.B, data = MathScoresAll, REML=TRUE)
```

Model A works (apparently), Model B gives warnings. They both give singular fits, but different components are singular ...

```{r diagnose}
## different log likelihoods (A is better)
logLik(math.lmerA)-logLik(math.lmerB)
singParams(math.lmerA)
singParams(math.lmerB)
```

## Refit with ML instead of REML

For comparison with `glmmTMB` fits (`glmmTMB` doesn't (yet) do REML), and out of curiosity.

```{r lme4_fit_ML,cache=TRUE}
math.lmerA.ML <- update(math.lmerA,REML=FALSE)
math.lmerB.ML <- update(math.lmerB,REML=FALSE)
```
Both models give warnings, although slightly different (3 vs 1 negative eigenvalues)

Now the log-likelihoods, singular parameters, and `VarCorr()` results are all the same (up to order) ...

```{r loglik_comp_ml}
logLik(math.lmerA.ML)-logLik(math.lmerB.ML) ## -2e-11
singParams(math.lmerA.ML)
singParams(math.lmerB.ML)
VarCorr(math.lmerA.ML)
VarCorr(math.lmerB.ML)
```

But the theta parameters themselves are **not** the same:
```{r compare_theta}
all.equal(sort(unname(th(math.lmerA.ML))),
          sort(unname(th(math.lmerB.ML))))
```
(Is this as expected? I thought there was a one-to-one mapping between $\theta$ and $\Sigma$?? FWIW the eigenvalues/vectors (`eigen(VarCorr(math.lmerA.ML)[[1]])`) are also identical (of course).)


More comparison:
```{r lFormula}
ff.A <- lFormula(form.A, data=MathScoresAll)
ff.B <- lFormula(form.B, data=MathScoresAll)
```

Z matrices are rearranged: model frames and fixed-effect X matrices are identical (except for attributes):

```{r show_Z}
gridExtra::grid.arrange(image(ff.A$reTrms$Zt[1:10,1:50]),
                        image(ff.B$reTrms$Zt[1:10,1:50]))
all.equal(c(ff.A$fr),c(ff.B$fr))
all.equal(c(ff.A$X),c(ff.B$X))
```


## comparing glmmTMB

```{r glmmTMB_fit,cache=TRUE}
math.tmbA <- glmmTMB(form.A, data = MathScoresAll)
math.tmbB <- glmmTMB(form.B, data = MathScoresAll)
```
Different warnings in each case ..

```{r glmmTMB_comp}
logLik(math.tmbA)-logLik(math.tmbB)
VarCorr(math.tmbA)
VarCorr(math.tmbB)
```

Nothing obviously fishy about the data ...

```{r ggpairs,cache=TRUE,warning=FALSE}
nm0 <- subset(MathScoresAll,select=-c(School,sizeScaled))
ggpairs(nm0,
        lower = list(continuous = wrap("points", size = 0.1)),
        progress=FALSE)
```

## To do

- Could instrument optimization functions, see the paths the optimization is taking (need a function to reorder theta parameters for consistency across models)



