---
title: "Covariance Structures lme4 vs glmmTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Autoscaling in lme4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

TODO:
- instead of a dataset, try to do it over simulated values instead for all comparisons...

## Introduction

`lme4` has four covariance classes/structures: `Covariance.us` (unstructured), `Covariance.diag` (diagonal), `Covariance.cs` (compound symmetry), and `Covarince.ar1` (autoregressive order 1). The syntax for use is somewhat similar to the `glmmTMB` package (see [covariance structures in glmmTMB](https://glmmtmb.github.io/glmmTMB/articles/covstruct.html)), although the results are slightly different.

Throughout the exploration, we will discuss the syntax differences.

## Reason for Computational Differences

When constructing generalized linear mixed models, `lme4` uses the Cholesky decomposition to reduce the problem to a function of the covariance parameters. In particular, `lme4` parameterizes the random-effects variance–covariance matrix $$
\Sigma_{\theta} = \sigma^{2} \Lambda_{\theta} \Lambda_{\theta}^{T},
$$ where the relative covariance factor (the Cholesky factor) $\Sigma_{\theta}$ is lower-triangular with positive diagonal entries, and is dependent on some variance-component parameter $\theta$. (For exact details of this structure, refer to the `lmer` [vignette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf).)

This allows optimization to proceed over the unconstrained elements of $\Sigma_{\theta}$, ensuring a gradient-free nonlinear optimization of the covariance parameter $\theta$'s.

Meanwhile, `glmmTMB` uses direct maximum likelihood estimation via Template Model Builder (TMB). Fixed effects ($\beta$'s), covariance parameters ($\theta$'s) and dispersion parameters are profiled out, and optimization proceeds on a likelihood parameterized by said fixed effects, covariance parameters, dispersion parameters, etc.

Despite these differences, we will show examples where `lme4` and `glmmTMB` provide similar estimates when they both use maximum likelihood estimation. By default, `lme4` uses the restricted maximum likelihood; hence in the following examples, we use `lmer(..., REML = FALSE)` to compare against `glmmTMB`.

## Comparison Setup

```{r setup}
library(lme4)
library(glmmTMB)

## Often want to ignore attributes and class.
all.equal.nocheck <- function(x, y, ..., check.attributes = FALSE, check.class = FALSE) {
  require("Matrix", quietly = TRUE)
  ## working around mode-matching headaches
  if (is(x, "Matrix")) x <- matrix(x)
  if (is(y, "Matrix")) y <- matrix(y)
  all.equal(x, y, ..., check.attributes = check.attributes, check.class = check.class)
}
```

To fit the linear mixed model, we will be using the classical `sleepstudy` dataset. Meanwhile, for the generalized linear mixed model, we will use the `Contraception` dataset from the `mlmRev` package:

```{r}
data(Contraception, package = "mlmRev")
```

## Unstructured (General Positive Definite)

This is the default setting for both `lme4` and `glmmTMB`. 
### Linear Mixed Model

```{r}
lme4.us <- lmer(Reaction ~ Days + us(Days | Subject), sleepstudy)
glmmTMB.us <- glmmTMB(Reaction ~ Days + us(Days | Subject), sleepstudy)

## Fixed effects
fixef(lme4.us); fixef(glmmTMB.us)$cond
all.equal.nocheck(fixef(lme4.us), fixef(glmmTMB.us)$cond)

## Sigma
sigma(lme4.us); sigma(glmmTMB.us)
all.equal.nocheck(sigma(lme4.us), sigma(glmmTMB.us))

## Log likelihoods
logLik(lme4.us); logLik(glmmTMB.us)
all.equal.nocheck(logLik(lme4.us), logLik(glmmTMB.us))
```

As expected, from here-on out, calculations related to the random-effects term differ.

```{r}
## Variance-Covariance Matrix
vcov(lme4.us); vcov(glmmTMB.us)$cond
all.equal.nocheck(vcov(lme4.us), vcov(glmmTMB.us)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(lme4.us)[[1]], 
          VarCorr(glmmTMB.us)$cond$Subject)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(lme4.us)$Subject, ranef(glmmTMB.us)$cond$Subject)
```

### Generalized Linear Mixed Model

```{r}
glme4.us <- glmer(use ~ age + urban + us(1 + urban | district), data = Contraception, 
                  family = binomial)

gglmmTMB.us <- glmmTMB(use ~ age + urban + us(1 + urban | district), data = Contraception, 
                       family = binomial)

## Fixed effects
fixef(glme4.us); fixef(gglmmTMB.us)$cond
all.equal.nocheck(fixef(glme4.us), fixef(gglmmTMB.us)$cond)

## Sigma
all.equal.nocheck(sigma(glme4.us), sigma(gglmmTMB.us))

## Log likelihoods
logLik(glme4.us); logLik(gglmmTMB.us)
all.equal.nocheck(logLik(glme4.us), logLik(gglmmTMB.us))
```

As expected, from here-on out, calculations related to the random-effects term differ.

```{r}
## Variance-Covariance Matrix
vcov(glme4.us); vcov(gglmmTMB.us)$cond
all.equal.nocheck(vcov(glme4.us), vcov(gglmmTMB.us)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(glme4.us)[[1]], 
          VarCorr(gglmmTMB.us)$cond$Subject)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(glme4.us)$district, ranef(gglmmTMB.us)$cond$district)
```

## Diagonal

The syntax is the same for fitting a heterogeneous diagonal covariance structure for `lme4` and `glmmTMB`. It differs when we want to fit a *homogeneous* diagonal covariance structure.

To demonstrate, if we wanted to fit a homogeneous diagonal covariance structure then we would write the following:

```{r, eval = FALSE}
lme4.us <- lmer(Reaction ~ Days + diag(Days | Subject, hom = TRUE), sleepstudy)
glmmTMB.us <- glmmTMB(Reaction ~ Days + homdiag(Days | Subject), sleepstudy)
```

Our focus will be the comparisons when we have a *heterogeneous* diagonal covariance structure as those cases are more interesting.

### Linear Mixed Model

```{r}
lme4.diag <- lmer(Reaction ~ Days + diag(Days | Subject), sleepstudy)
glmmTMB.diag <- glmmTMB(Reaction ~ Days + diag(Days | Subject), sleepstudy)

## Fixed effects
fixef(lme4.diag); fixef(glmmTMB.diag)$cond
all.equal.nocheck(fixef(lme4.diag), fixef(glmmTMB.diag)$cond)

## Sigma
sigma(lme4.diag); sigma(glmmTMB.diag)
all.equal.nocheck(sigma(lme4.diag), sigma(glmmTMB.diag))

## Log likelihoods
logLik(lme4.diag); logLik(glmmTMB.diag)
all.equal.nocheck(logLik(lme4.diag), logLik(glmmTMB.diag))

## Variance-Covariance Matrix
vcov(lme4.diag); vcov(glmmTMB.diag)$cond
all.equal.nocheck(vcov(lme4.diag), vcov(glmmTMB.diag)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(lme4.diag)[[1]], 
          VarCorr(glmmTMB.diag)$cond$Subject)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(lme4.diag)$Subject, ranef(glmmTMB.diag)$cond$Subject)
```

### Generalized Linear Mixed Model

```{r}
glme4.diag <- glmer(use ~ age + urban + diag(1 + urban | district), data = Contraception, 
                  family = binomial)

gglmmTMB.diag <- glmmTMB(use ~ age + urban + diag(1 + urban | district), data = Contraception, 
                       family = binomial)

## Fixed effects
fixef(glme4.diag); fixef(gglmmTMB.diag)$cond
all.equal.nocheck(fixef(glme4.diag), fixef(gglmmTMB.diag)$cond)

## Sigma
all.equal.nocheck(sigma(glme4.diag), sigma(gglmmTMB.diag))

## Log likelihoods
logLik(glme4.diag); logLik(gglmmTMB.diag)
all.equal.nocheck(logLik(glme4.diag), logLik(gglmmTMB.diag))

## Variance-Covariance Matrix
vcov(glme4.diag); vcov(gglmmTMB.diag)$cond
all.equal.nocheck(vcov(glme4.diag), vcov(gglmmTMB.diag)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(glme4.diag)[[1]], 
          VarCorr(gglmmTMB.diag)$cond$district)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(glme4.diag)$district, ranef(gglmmTMB.diag)$cond$district)
```

## Compound Symmetry

Similar to the diagonal case, the syntax is the same for fitting a heterogeneous compound symmetry covariance structure for `lme4` and `glmmTMB`. Again, it differs when we want to fit a *homogeneous* compound symmetry covariance structure.

To demonstrate, if we wanted to fit a homogeneous compound symmetry covariance structure then we would write the following:

```{r, eval = FALSE}
lme4.us <- lmer(Reaction ~ Days + cs(Days | Subject, hom = TRUE), sleepstudy)
glmmTMB.us <- glmmTMB(Reaction ~ Days + homcs(Days | Subject), sleepstudy)
```

Once more, we will make comparisons against a *heterogeneous* compound symmetric covariance structure.

### Linear Mixed Model

```{r}
lme4.cs <- lmer(Reaction ~ Days + cs(Days | Subject), sleepstudy)
glmmTMB.cs <- glmmTMB(Reaction ~ Days + cs(Days | Subject), sleepstudy)

## Fixed effects
fixef(lme4.cs); fixef(glmmTMB.cs)$cond
all.equal.nocheck(fixef(lme4.cs), fixef(glmmTMB.cs)$cond)

## Sigma
sigma(lme4.cs); sigma(glmmTMB.cs)
all.equal.nocheck(sigma(lme4.cs), sigma(glmmTMB.cs))

## Log likelihoods
logLik(lme4.cs); logLik(glmmTMB.cs)
all.equal.nocheck(logLik(lme4.cs), logLik(glmmTMB.cs))

## Variance-Covariance Matrix
vcov(lme4.cs); vcov(glmmTMB.cs)$cond
all.equal.nocheck(vcov(lme4.cs), vcov(glmmTMB.cs)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(lme4.cs)[[1]], 
          VarCorr(glmmTMB.cs)$cond$Subject)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(lme4.cs)$Subject, ranef(glmmTMB.cs)$cond$Subject)
```

### Generalized Linear Mixed Model

```{r}
glme4.cs <- glmer(use ~ age + urban + cs(1 + urban | district), data = Contraception, 
                  family = binomial)

gglmmTMB.cs <- glmmTMB(use ~ age + urban + cs(1 + urban | district), data = Contraception, 
                       family = binomial)

## Fixed effects
fixef(glme4.cs); fixef(gglmmTMB.cs)$cond
all.equal.nocheck(fixef(glme4.cs), fixef(gglmmTMB.cs)$cond)

## Sigma
all.equal.nocheck(sigma(glme4.cs), sigma(gglmmTMB.cs))

## Log likelihoods
logLik(glme4.cs); logLik(gglmmTMB.cs)
all.equal.nocheck(logLik(glme4.cs), logLik(gglmmTMB.cs))

## Variance-Covariance Matrix
vcov(glme4.cs); vcov(gglmmTMB.cs)$cond
all.equal.nocheck(vcov(glme4.cs), vcov(gglmmTMB.cs)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(glme4.cs)[[1]], 
          VarCorr(gglmmTMB.cs)$cond$district)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(glme4.cs)$district, ranef(gglmmTMB.cs)$cond$district)
```

## Autoregressive Order 1

For this comparison, we will focus on a simulated data set where we explicitly denote $\rho = 0.7$ to better compare the results between the two models. 

```{r}
set.seed(1)
simGroup <- function(g, n=6, phi=0.7) {
  x <- MASS::mvrnorm(mu = rep(0,n),
                     Sigma = phi^as.matrix(dist(1:n)) )  
  y <- x + rnorm(n)                              
  times <- factor(1:n)
  group <- factor(rep(g,n))
  data.frame(y, times, group)
}
dat1 <- do.call("rbind", lapply(1:5000, simGroup))
```

Unlike the diagonal and compound symmetry case, the syntax is different for fitting both a heterogeneous and homogeneous ar1 model for `lme4` and `glmmTMB`.

For a *heterogeneous* ar1 covariance structure we would write the following:

```{r, eval = FALSE}
lme4.ar1 <- lmer(y ~ times + ar1(0 + times | group), data = dat1, REML = FALSE)
glmmTMB.ar1 <- glmmTMB(y ~ times + ar1(0 + times | group), data = dat1)
```

We will instead focus on comparisons for a *homogeneous* ar1 covariance structure.

```{r}
lme4.ar1 <- lmer(y ~ times + ar1(0 + times | group, hom = TRUE), data = dat1, REML = FALSE)
glmmTMB.ar1 <- glmmTMB(y ~ times + hetar1(0 + times | group), data = dat1)

## Fixed effects
fixef(lme4.ar1); fixef(glmmTMB.ar1)$cond
all.equal.nocheck(fixef(lme4.ar1), fixef(glmmTMB.ar1)$cond)

## Sigma
sigma(lme4.ar1); sigma(glmmTMB.ar1)
all.equal.nocheck(sigma(lme4.ar1), sigma(glmmTMB.ar1))

## Log likelihoods
logLik(lme4.ar1); logLik(glmmTMB.ar1)
all.equal.nocheck(logLik(lme4.ar1), logLik(glmmTMB.ar1))

## Variance-Covariance Matrix
all.equal.nocheck(vcov(lme4.ar1), vcov(glmmTMB.ar1)$cond)

## Variance and Covariance Components
all.equal.nocheck(VarCorr(lme4.ar1)[[1]], 
          VarCorr(glmmTMB.ar1)$cond$group)

## Conditional Modes of the Random Effects
all.equal.nocheck(ranef(lme4.ar1)$group, ranef(glmmTMB.ar1)$cond$group)
```



