---
title: "Downstream methods"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Downstream methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

There are many methods from other packages that are compatible with `merMod` objects that others may find useful for plot diagnostics, model comparisons, visual plotting, and regression tables. This vignette provides instructions on some recommended packages with some examples, although this list is far from exhaustive. We aimed to include the more popular packages.

For questions regarding how to use (generalized) linear mixed models, see this [FAQ.](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#introduction)

# Diagnostic Plots

While diagnostic plots can be evoked by using `plot(merMod_object)` (see `??lme4:plot.merMod`), some of these plots may be supplemented or superseded by functions from other packages. 

This section emphasizes the `performance` and `DHARMa` package, although there are many other examples such as `lattice::qqmath()`.

```{r}
#| message: false
#| warning: false
library(lme4)
# Example of a linear mixed-effects model (LMM)
mod_ss <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
# Assume a non-mixed model for comparison
mod_ss2 <- lm(Reaction ~ Days, sleepstudy)

# Another (LMM) example, and others for comparison
mod_cw = lmer(weight ~ Time + Diet + (1|Chick), data = ChickWeight, REML = FALSE)
mod_cw2 = lm(weight ~ Time + Diet, data = ChickWeight)
mod_cw3 = lmer(weight ~ Time + (1|Chick), data = ChickWeight, REML = FALSE)

# Example of a Poisson generalized linear mixed-effects model (GLMM)
form <- TICKS~YEAR+HEIGHT+(1|BROOD)+(1|INDEX)+(1|LOCATION)
mod_gt  <- glmer(form, family="poisson",data=grouseticks)
```

## Performance

The [performance](https://easystats.github.io/performance/) package is excellent for evaluating the quality of model fit and works well with `merMod` objects (and many others). The first example we'll show is `performance::check_model()`, which does checks for classical assumptions used for most linear models: normality of residuals, linear relationship, homogeneity of variance, outliers. 

For mixed-effect models, it also checks for normality of random effects. The uncommonly used posterior predictive checks is meant to see whether the distributional family used fits well to the data.

```{r, eval = require('performance'), message=FALSE, warning=FALSE, fig.width=8, fig.height=6, dpi=96}
performance::check_model(mod_ss)
```

For those working with mixed models, it may important to check whether some dimensions of the variance-covariance are estimated to be zero.

```{r, eval = require('performance')}
performance::check_singularity(mod_ss)
```

## DHARMa

The [DHARMa](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html) package uses a simulation-based method for residual checks for generalized linear (mixed) models. 

```{r, eval = require('DHARMa'), message=FALSE, warning=FALSE, fig.width=7, fig.height=5, dpi=96}
simulationOutput <- DHARMa::simulateResiduals(fittedModel = mod_ss)
plot(simulationOutput)
```

The left plot performs three tests, the Kolmogorov–Smirnov test, simulation-based dispersion tests (see `??DHARMa::testDispersion`), and simulation-based outlier tests (see `??DHARMa::testOutliers`).

## Car 

The [car](https://CRAN.R-project.org/package=car) package was designed for "An R Companion to Applied Regression" to provide functions that are applied to a fitted regression model. The `car::influencePlot()` creates a diagnostic bubble plot that visualizes influential observations in a regression by plotting Studentized residuals against leverage (hat values), with bubble size representing Cook's distance.

```{r, eval = require('car'), message=FALSE, warning=FALSE, fig.width=7, fig.height=5, dpi=96}
car::influencePlot(mod_ss)
```

# Model Comparison & Performance

Most model comparisons for `merMod` objects will focus on the fixed effects, and the ones shown will be just for fixed effects unless otherwise stated.

## Performance

```{r, eval = require('performance')}
performance::model_performance(mod_ss)
```

We can also use `performance::compare_performance()`.

```{r, eval = require('performance')}
performance::compare_performance(mod_ss, mod_ss2)
```

## Pbkrtest

The [pbkrtest](https://hojsgaard.github.io/pbkrtest/index.html) package implements three tests to examine fixed effects from linear mixed-effects models (LMM):
- Parametric bootstrap test `pbkrtest::PBmodcomp()`
- Kenward-Roger-type F-test `pbkrtest::KRmodcomp()`
- Satterthwaite-type F-test `pbkrtest::SATmodcomp()`

```{r, eval = require('pbkrtest')}
fm0 <- lmer(weight ~ Time + (1|Chick), data = ChickWeight)
fm1 <- update(fm0, .~.-Time)

pbkrtest::PBmodcomp(fm0, fm1)
pbkrtest::KRmodcomp(fm0, fm1)
pbkrtest::SATmodcomp(fm0, fm1)
```

Note: the `pbkrtest::KRmodcomp()` function was package was intended for `lmerMod` objects only, as the Kenward-Roger-type F-test was designed for fitted linear mixed models.

## MuMIn

The [MuMIn](https://cran.r-universe.dev/MuMIn/doc/manual.html) package provides tools for model selection by automating the process through subsets of the maximum model.

```{r, eval = require('MuMIn'), message = FALSE, warning = FALSE}
options(na.action = "na.fail")

mumin_compare = MuMIn::dredge(mod_cw)
mumin_compare
```

`MuMIn::dredge()` doesn't print the fitted model object, so to find which the labelled models 1, 2, etc., correspond to, use `get.models(mumin_compare, subset = TRUE)`. The output tends to be long and is omitted for space reasons. 

## RLRsim

A common way to test the significance of random effects is through a likelihood ratio test. The [RLRsim](https://github.com/fabian-s/RLRsim) package includes fast simulation-based exact tests that is compatible with `lmerMod` objects.

```{r, eval = require('RLRsim')}
# First model is one with the random effects,
# the second is must be a lm-object.
RLRsim::exactLRT(m = mod_cw, m0 = mod_cw2)
```

## Multcomp

The [multcomp](https://multcomp.r-forge.r-project.org/) package was designed for performing general linear hypotheses in parametric models.

```{r, eval = require('multcomp')}
glh_cw = multcomp::glht(mod_cw)
summary(glh_cw)
```

We can also look at confidence intervals for the pairwise difference between groups:

```{r, eval = require('multcomp'), fig.width=6.5, fig.height=4, dpi=100}
plot(glh_cw)
```

## Car

`car::Anova()` can be used traditional ANOVA-style tables using Wald $\chi^{2}$ statistics on the fixed effects only.

```{r, eval = require('car')}
car::Anova(mod_cw)
```

## Insight

As its name suggests, the [insight](https://easystats.github.io/insight/index.html) package was designed to easily gather information from statistical models from a variety of different object types. Their documentation already covers examples of `merMod` objects [here](https://easystats.github.io/insight/articles/insight.html#examples).

# Plotting & Prediction

When interpreting results, we may be interested in seeing the coefficient plots or effect plots. Usually, the first step requires predicting the response and then fitting the model (using `plot()` or `ggplot()`).

```{r, eval=require('ggplot2')}
library(ggplot2)
```

## Dotwhisker

The [dotwhisker](https://fsolt.org/dotwhisker/) package was designed to easily make dot-and-whisker plots from regression results. We recommend scaling by two standard deviations as recommended by [Gelman (2008).](https://doi.org/10.1002/sim.3107).

```{r, eval=require('dotwhisker'), message=FALSE, warning=FALSE, fig.width=4, fig.height=2.4, dpi=150}

effects_gt <- broom.mixed::tidy(mod_cw, effect = "fixed", conf.int = TRUE)

dotwhisker::dwplot(effects_gt, by_2sd = TRUE)
```

Notice that we combined the usage of [broom.mixed](https://github.com/bbolker/broom.mixed) package to extract the fixed effects (for the dot-and-whisker plots). Details can be found by clicking the url for broom.mixed or see [this section.](#broom-and-broom.mixed)

## Emmeans

As its name implies, the [emmeans](https://rvlenth.github.io/emmeans/) package obtains the estimated marginal means.

More details and examples for `lmerMod` objects can be found [here.](https://cran.r-project.org/web/packages/emmeans/vignettes/sophisticated.html)
```{r, eval = require('emmeans'), message=FALSE, warning=FALSE, fig.width=4, fig.height=2.4, dpi=150}
plot(emmeans::emmeans(mod_cw, "Diet"))
```

## SjPlot

The [sjPlot](https://strengejacke.github.io/sjPlot/) package was built for easier plotting and table outputs. 

```{r, eval = require('sjPlot'), message=FALSE, warning=FALSE, fig.width=4, fig.height=2.4, dpi=150}
sjPlot::plot_model(mod_cw)
```

## ggEffects

The [ggEffects](https://strengejacke.github.io/ggeffects/) package also computes the marginal effects, returning ready-to-graph results via the [ggplot2](https://ggplot2.tidyverse.org/) package.

The [ggEffects documentation](https://strengejacke.github.io/ggeffects/articles/introduction_randomeffects.html) already includes more details and examples for `lmerMod` objects.

```{r, eval = require('ggEffects'), message=FALSE, warning=FALSE, fig.width=4, fig.height=2.4, dpi=150}
predict_ss <- ggeffects::predict_response(mod_ss, "Days [0:9]")
ggplot(predict_ss, aes(x, predicted)) +
  geom_line() + labs(x = "Days", y = "Predicted Reaction") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.1)
```

The ggEffects package is supposed to be superseded by [modelbased](https://easystats.github.io/modelbased/), however, it may run errors when using `lmerMod` objects.

## Marginaleffects

As its package name implies, the goal of [marginaleffects](https://marginaleffects.com/) is to compute the marginal means as well as various other model comparisons for a lot of different classes, `merMod` objects being one of them.

More details and examples for mixed models can be found [here.](https://marginaleffects.com/bonus/lme4.html)

```{r, eval = require('marginaleffects'),message=FALSE, warning=FALSE, fig.width=4, fig.height=2.4, dpi=150}
library(marginaleffects)
# Note: datagrid depends on marginaleffects

pred_ss <- predictions(mod_ss, newdata = datagrid(Days = 0:9))
# Plot using ggplot2
ggplot(pred_ss, aes(x = Days, y = estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.1) +
  labs(x = "Days", y = "Predicted Reaction")
```

# Formatted Regression Tables

Although the output can easily be seen by applying `summary()` to the `merMod` object, it can be useful to use packages that automatically format the results nicely into $\LaTeX$, HTML, and text output. There are many such functions that appear to be compatiable with `merMod` objects.

In this section, we display the code (though it is not executed) along with the resulting HTML output. The corresponding versions for $\LaTeX$ and plain text are equally straightforward.

## SjPlot

The [sjPlot](https://strengejacke.github.io/sjPlot/) package can be used for table outputs in HTML. There is a fantastic example written in their documentation [here.](https://strengejacke.github.io/sjPlot/articles/tab_model_robust.html?q=lme4#robust-covariance-matrix-estimation-for-mixed-models)

## Parameters

The [Parameters](https://easystats.github.io/parameters/index.html) package, as its name suggests, extracts parameters from a fitted model. Their own documentation also includes examples of mixed models from lme4 [here.](https://easystats.github.io/parameters/index.html) and exact table formatting can be found [here.](https://easystats.github.io/parameters/articles/model_parameters_formatting.html)

## Huxtable

The [Huxtable](https://hughjonesd.github.io/huxtable/) package can be used to create $\LaTeX$ (see `huxtable::print_latex()`) and HTML (see `huxtable::print_html()`) output.  

```{r, eval = FALSE}
h_tab = huxtable::huxreg(mod_ss)
huxtable::print_html(h_tab) 
```

<table class="huxtable" data-quarto-disable-processing="true" style="border-collapse: collapse; border: 0px; margin-bottom: 2em; margin-top: 2em; ; margin-left: auto; margin-right: auto;  ">
<col><col><tr>
<th style="vertical-align: top; text-align: center; white-space: normal; border-style: solid solid solid solid; border-width: 0.8pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><th style="vertical-align: top; text-align: center; white-space: normal; border-style: solid solid solid solid; border-width: 0.8pt 0pt 0.4pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(1)</th></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(Intercept)</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0.4pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">251.405&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(6.825)</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">Days</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">10.467&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(1.546)</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">sd__(Intercept)</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">24.741&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(NA)&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">cor__(Intercept).Days</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">0.066&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(NA)&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">sd__Days</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">5.922&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(NA)&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">sd__Observation</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">25.592&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"></th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0.4pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">(NA)&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">N</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0.4pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">180&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">logLik</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">-871.814&nbsp;</td></tr>
<tr>
<th style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0.8pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">AIC</th><td style="vertical-align: top; text-align: right; white-space: normal; border-style: solid solid solid solid; border-width: 0pt 0pt 0.8pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;">1755.628&nbsp;</td></tr>
<tr>
<th colspan="2" style="vertical-align: top; text-align: left; white-space: normal; border-style: solid solid solid solid; border-width: 0.8pt 0pt 0pt 0pt;    padding: 6pt 6pt 6pt 6pt; font-weight: normal;"> *** p &lt; 0.001;  ** p &lt; 0.01;  * p &lt; 0.05.</th></tr>
</table>

## Stargazer

[Stargazer](https://cran.r-project.org/web/packages/stargazer/index.html) was intended to create well formatted regression tables in $\LaTeX$, HTML/CSS, and plain text. 

```{r, eval = FALSE}
stargazer::stargazer(mod_ss, type = "html") # use type = "latex" for LaTeX
```

<table style="text-align:center"><tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"></td><td><em>Dependent variable:</em></td></tr>
<tr><td></td><td colspan="1" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left"></td><td>Reaction</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">Days</td><td>10.467<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td>(1.546)</td></tr>
<tr><td style="text-align:left"></td><td></td></tr>
<tr><td style="text-align:left">Constant</td><td>251.405<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td>(6.825)</td></tr>
<tr><td style="text-align:left"></td><td></td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">Observations</td><td>180</td></tr>
<tr><td style="text-align:left">Log Likelihood</td><td>-871.814</td></tr>
<tr><td style="text-align:left">Akaike Inf. Crit.</td><td>1,755.628</td></tr>
<tr><td style="text-align:left">Bayesian Inf. Crit.</td><td>1,774.786</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"><em>Note:</em></td><td style="text-align:right"><sup>*</sup>p<0.1; <sup>**</sup>p<0.05; <sup>***</sup>p<0.01</td></tr>
</table>

## Texreg

We could also use the [texreg](https://leifeld.r-universe.dev/texreg) package to create $\LaTeX$ `texreg::texreg()` and HTML `texreg::htmlreg()` tables.

```{r, eval = FALSE}
texreg::htmlreg(mod_ss)
```

<table class="texreg" style="margin: 10px auto;border-collapse: collapse;border-spacing: 0px;caption-side: bottom;color: #000000;border-top: 2px solid #000000;">
<caption>Statistical models</caption>
<thead>
<tr>
<th style="padding-left: 5px;padding-right: 5px;">&nbsp;</th>
<th style="padding-left: 5px;padding-right: 5px;">Model 1</th>
</tr>
</thead>
<tbody>
<tr style="border-top: 1px solid #000000;">
<td style="padding-left: 5px;padding-right: 5px;">(Intercept)</td>
<td style="padding-left: 5px;padding-right: 5px;">251.41<sup>&#42;&#42;&#42;</sup></td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">&nbsp;</td>
<td style="padding-left: 5px;padding-right: 5px;">(6.82)</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Days</td>
<td style="padding-left: 5px;padding-right: 5px;">10.47<sup>&#42;&#42;&#42;</sup></td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">&nbsp;</td>
<td style="padding-left: 5px;padding-right: 5px;">(1.55)</td>
</tr>
<tr style="border-top: 1px solid #000000;">
<td style="padding-left: 5px;padding-right: 5px;">AIC</td>
<td style="padding-left: 5px;padding-right: 5px;">1755.63</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">BIC</td>
<td style="padding-left: 5px;padding-right: 5px;">1774.79</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Log Likelihood</td>
<td style="padding-left: 5px;padding-right: 5px;">-871.81</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Num. obs.</td>
<td style="padding-left: 5px;padding-right: 5px;">180</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Num. groups: Subject</td>
<td style="padding-left: 5px;padding-right: 5px;">18</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Var: Subject (Intercept)</td>
<td style="padding-left: 5px;padding-right: 5px;">612.10</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Var: Subject Days</td>
<td style="padding-left: 5px;padding-right: 5px;">35.07</td>
</tr>
<tr>
<td style="padding-left: 5px;padding-right: 5px;">Cov: Subject (Intercept) Days</td>
<td style="padding-left: 5px;padding-right: 5px;">9.60</td>
</tr>
<tr style="border-bottom: 2px solid #000000;">
<td style="padding-left: 5px;padding-right: 5px;">Var: Residual</td>
<td style="padding-left: 5px;padding-right: 5px;">654.94</td>
</tr>
</tbody>
<tfoot>
<tr>
<td style="font-size: 0.8em;" colspan="2"><sup>&#42;&#42;&#42;</sup>p &lt; 0.001; <sup>&#42;&#42;</sup>p &lt; 0.01; <sup>&#42;</sup>p &lt; 0.05</td>
</tr>
</tfoot>
</table>

## Memisc

The goal of the [memisc](https://melff.github.io/memisc/) package is to make it easier to deal with survey data sets, such as table formatting in $\LaTeX$ (using `memisc::mtable_format_latex()`) and HTML (using `memisc::mtable_format_html()`).

```{r, eval = FALSE}
mem_tab <- memisc::mtable("Model 1"=mod_cw,"Model 2"=mod_cw2,
                          "Model 3"=mod_cw3, summary.stats=c("sigma","R-squared","F","p","N"))

memisc::mtable_format_html(mem_tab)
```

<style>
      table#mtable-3 {
         border-collapse: collapse; border-style: none; margin: 2ex auto;
      } 
      table#mtable-3 tr {
          border-style: none;
      } 
      table#mtable-3 td {
        padding-top: 1px; 
        padding-bottom: 1px; 
        padding-left: 0.5ex; 
        padding-right: 0.5ex; 
        margin: 0px; 
        margin-top: 0px;
        margin-bottom: 0px;
        border-style: none; 
        border-width: 0px; 
      }
table#mtable-3 tr:nth-child(1){
     border-top: 2px solid;
}

table#mtable-3 tr:nth-child(1),
table#mtable-3 tr:nth-child(11){
     border-bottom: 1px solid;
}
table#mtable-3 tr:nth-child(16){
     border-bottom: 2px solid;
} table#mtable-3 td:nth-child(3),
table#mtable-3 td:nth-child(6),
table#mtable-3 td:nth-child(9){
         padding-left: 0px;
         padding-right: 0px;
         text-align: center; 
         /*background-color: blue;*/
        } table#mtable-3 td:nth-child(2),
table#mtable-3 td:nth-child(5),
table#mtable-3 td:nth-child(8){
         padding-left: 0.5ex;
         padding-right: 0px;
         text-align: right; 
         /*background-color: red;*/
        } table#mtable-3 td:nth-child(4),
table#mtable-3 td:nth-child(7),
table#mtable-3 td:nth-child(10){
         padding-left: 0px;
         padding-right: 0.5ex;
         text-align: left; 
         /*background-color: green;*/
        } 
table#mtable-3 td:nth-child(n of .header){
         padding-left: 0.5ex;
         padding-right: 0.5ex;
         text-align: center; 
         /*background-color: gray;*/
        } 
</style><table id="mtable-3">
<tr><td colspan="1" class="header"></td><td colspan="3" class="header">Model 1</td><td colspan="3" class="header">Model 2</td><td colspan="3" class="header">Model 3</td></tr>
<tr><td>(Intercept)</td><td>11</td><td>.</td><td>244</td><td>10</td><td>.</td><td>924<span class="signif.symbol">**</span></td><td>27</td><td>.</td><td>845<span class="signif.symbol">***</span></td></tr>
<tr><td></td><td>(5</td><td>.</td><td>789)</td><td>(3</td><td>.</td><td>361)</td><td>(4</td><td>.</td><td>388)</td></tr>
<tr><td>Time</td><td>8</td><td>.</td><td>717<span class="signif.symbol">***</span></td><td>8</td><td>.</td><td>750<span class="signif.symbol">***</span></td><td>8</td><td>.</td><td>726<span class="signif.symbol">***</span></td></tr>
<tr><td></td><td>(0</td><td>.</td><td>175)</td><td>(0</td><td>.</td><td>222)</td><td>(0</td><td>.</td><td>176)</td></tr>
<tr><td>Diet: 2/1</td><td>16</td><td>.</td><td>210</td><td>16</td><td>.</td><td>166<span class="signif.symbol">***</span></td><td></td><td></td><td></td></tr>
<tr><td></td><td>(9</td><td>.</td><td>464)</td><td>(4</td><td>.</td><td>086)</td><td></td><td></td><td></td></tr>
<tr><td>Diet: 3/1</td><td>36</td><td>.</td><td>543<span class="signif.symbol">***</span></td><td>36</td><td>.</td><td>499<span class="signif.symbol">***</span></td><td></td><td></td><td></td></tr>
<tr><td></td><td>(9</td><td>.</td><td>464)</td><td>(4</td><td>.</td><td>086)</td><td></td><td></td><td></td></tr>
<tr><td>Diet: 4/1</td><td>30</td><td>.</td><td>013<span class="signif.symbol">**</span></td><td>30</td><td>.</td><td>233<span class="signif.symbol">***</span></td><td></td><td></td><td></td></tr>
<tr><td></td><td>(9</td><td>.</td><td>471)</td><td>(4</td><td>.</td><td>107)</td><td></td><td></td><td></td></tr>
<tr><td>N</td><td>578</td><td></td><td></td><td>578</td><td></td><td></td><td>578</td><td></td><td></td></tr>
<tr><td>R-squared</td><td></td><td></td><td></td><td>0</td><td>.</td><td>745</td><td></td><td></td><td></td></tr>
<tr><td>sigma</td><td></td><td></td><td></td><td>35</td><td>.</td><td>993</td><td></td><td></td><td></td></tr>
<tr><td>F</td><td></td><td></td><td></td><td>419</td><td>.</td><td>177</td><td></td><td></td><td></td></tr>
<tr><td>p</td><td></td><td></td><td></td><td>0</td><td>.</td><td>000</td><td></td><td></td><td></td></tr>
<tr><td colspan="10"><p>Significance: *** = p &lt; 0.001; ** = p &lt; 0.01; * = p &lt; 0.05</p>
</td></tr>
</table>

# Other

This section covers other external packages that may be useful but does not fall under the category of previous sections.

## Broom.mixed

The [broom](https://broom.tidymodels.org/) package takes output from built-in functions in R and converts them into a [tibble](https://tibble.tidyverse.org/), which is an alternative to R's built in `data.frame`.

[Broom.mixed](https://cran.r-project.org/web/packages/broom.mixed/index.html) is a spin-off of broom where it takes outputs from outputs from specifically mixed models from various popular mixed model packages, and of course `lme4` is one of them.

```{r, eval = require('broom.mixed')}
broom.mixed::tidy(mod_ss)
```

## Equatiomatic

The [equatiomatic](https://datalorax.github.io/equatiomatic/) package takes a fitted model and writes the equation for you, formatted in $\LaTeX$.

```{r, eval = require('equatiomatic')}
#| eval: false
equatiomatic::extract_eq(mod_ss)
```

The output provides $\LaTeX$ code that outputs the following equation:
$$
\begin{aligned}
  \operatorname{Reaction}_{i}  &\sim N \left(\alpha_{j[i]} + \beta_{1j[i]}(\operatorname{Days}), \sigma^2 \right) \\    
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\alpha_{j} \\
      &\beta_{1j}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\mu_{\alpha_{j}} \\
      &\mu_{\beta_{1j}}
    \end{aligned}
  \end{array}
\right)
, 
\left(
  \begin{array}{cc}
     \sigma^2_{\alpha_{j}} & \rho_{\alpha_{j}\beta_{1j}} \\ 
     \rho_{\beta_{1j}\alpha_{j}} & \sigma^2_{\beta_{1j}}
  \end{array}
\right)
 \right)
    \text{, for Subject j = 1,} \dots \text{,J}
\end{aligned}
$$


