## Laplace diagnostics

Copied/modified from DMB's code in glmer paper draft ...

```{r pkgs,message=FALSE}
library(lme4)
library(lattice)
library(ggplot2); theme_set(theme_bw())
library(plyr)      ## for reshaping
library(abind)     ## ditto
library(reshape2)  ## for melt() generic
```

For now, I'm going to leave this function echoed in all its glory ...
```{r zetafun}
zeta <- function(m, zmin=-3, zmax=3, npts=NULL,
                 grps=NULL) {
    stopifnot (is(m, "glmerMod"),
               length(m@flist) == 1L)    # single grouping factor
    nvar <- length(m@cnms[[1]])
    if (nvar>2) stop("can't handle vector RE with length >2")
    if (is.null(npts)) npts <- if (nvar>1) 31L else 301L
    pp <- m  ## ??? why copy? (something to do with reference classes?
             ## but this is still not a deep copy ...
    rr <- m@resp                         ## extract response module
    u0 <- getME(pp,"u")                  ## conditional modes
    L <- getME(pp,"L")
    ## sd <- 1/getME(pp,"L")@x
    ## filled elements of L matrix==diag for simple case
    ## for more general case need the following -- still efficient
    sd <- sqrt(diag(chol2inv(L)))
    ff <- getME(pp,"flist")[[1]]
    if (is.null(grps)) grps <- seq(length(levels(ff)))
    ngrps <- length(grps)
    ## fixed-effects contribution to linear predictor
    fc <- getME(pp,"X") %*% getME(pp,"beta") 
    ZL <- t(getME(pp,"Lambdat") %*% getME(pp,"Zt"))
    ## evaluate the unscaled conditional density on the deviance scale
    dc <- function(z) {
        uu <- u0 + z * sd    ## displace conditional modes
        ##  should still work if z is a vector (by recycling, because u values
        ##  applying to each group are stored adjacent to each other)
        rr$updateMu(fc + ZL %*% uu)     ## update linear predictor
        drc <- unname(as.vector(tapply(rr$devResid(), ff, sum)))
        uuc <- colSums(matrix(uu * uu,nrow=nvar))
        (drc + uuc)[grps]
    }
    zvals <- seq(zmin, zmax, length.out = npts)
    if (nvar==1) { # scalar-valued random effects
        vv <- vapply(zvals,dc,numeric(ngrps), USE.NAMES=FALSE)
        vv <- t(vv)  ## n.z * n.id
    } else { # vector-valued random effects
        nz <- length(zvals)
        vv <- mapply(function(x,y) { dc(c(x,y)) },
                         rep(zvals,nz),rep(zvals,each=nz))
        ## result: nu*(nz^2) matrix; want nz*nz*nu array
        ## *with* each nu slice being a nz^2 matrix for one group
        ## I'm sure there's a clever way to do this with array/aperm,
        ## but I just couldn't figure it out.  Instead,
        ## (1) take each row of vv and make into a matrix, return as list
        ##     of matrices
        ## (2) bind matrices into an array
        vv <- do.call(abind,c(alply(vv,1,matrix,nrow=nz),list(along=3)))
    }
    d0 <- dc(0) # because this is the last evaluation, the model is restored to its incoming state
    sweep.margin <- if (nvar==1) 2 else 3 
    devarr <- sweep(vv,sweep.margin,d0,"-")
    ## computing deviance rather than signed sqrt, since we're not using it
    ## anyway and it's harder to generalize to >1 dimension ...
    rr <- list(zvals=zvals,
               devarr=devarr)
    ## signed square root
    ## array(ifelse(zvals < 0, -1, 1), c(npts, length(u0))))
    class(rr) <- "laplaceDiag"
    rr
}
```

Not shown: `melt` function for these objects that converts them from a list ($z$ value vector plus array of deviances) to a data frame ...
```{r echo=FALSE}
melt.laplaceDiag <- function(data,...) {
    require(reshape2)
    zvals <- data$zval
    if (length(dim(data$devarr))==2) {
        n.id <- ncol(data$devarr)
        n.z <- nrow(data$devarr)
        data.frame(id=gl(n.id,n.z),
                   zvals,
                   dev=c(data$devarr))
    } else {
        ## assume for now same z resolution for both dimensions
        n.z <- dim(data$devarr)[2]
        n.id <- dim(data$devarr)[3]
        data.frame(id=gl(n.id,n.z^2),
                   zval1=rep(zvals,n.z),  ## recycle
                   zval2=rep(zvals,each=n.z), ## recycle
                   dev=c(data$devarr))
    }
}

dnorm2d <- function(x) {
    dnorm(x)/sqrt(2*pi)  ## == exp(-x^2/2)/(2*pi)
}

dnorm2d2 <- function(z1,z2) {
    exp(-(z1^2+z2^2)/2)/(2*pi)
}

plot.laplaceDiag <- function(x,scaled=FALSE,
                             type=c("g","l"),
                             aspect=0.6,
                             xlab="z",ylab="density",
                             ...) {
    nvar <- length(dim(x$devarr))-1
    mm <- melt(x)
    mm <- transform(mm,
                    y = if (nvar==1) {
                        if (!scaled) {
                            dnorm(sqrt(dev))
                        } else {
                            dnorm(sqrt(dev))/dnorm(zvals)
                        }
                    } else {
                        if (!scaled) {
                            dnorm2d(sqrt(dev))
                        } else {
                            dnorm2d(sqrt(dev))/dnorm2d2(zval1,zval2)
                        }
                    })
    if (nvar==1) {
        print(xyplot(y ~ zvals|id, data=mm,
               type=type, aspect=aspect,
               xlab=xlab,ylab=ylab,
               ...,
               panel=function(x,y,...){
                   if (!scaled) {
                       panel.lines(x, dnorm(x), lty=2)
                   } else {
                       panel.abline(h=1, lty=2)
                   }
                   panel.xyplot(x,y,...)
               }))
    } else {
        print(contourplot(y ~ zval1*zval2|id, data=mm,
                    type=type, aspect=aspect,
                    labels=FALSE,
                    xlab=xlab,ylab=ylab,
                    scales=list(z=list(relation="free"))))
    }
    invisible(mm)
}
```

Replicate glmer paper Figs 2/3:

```{r fit1,cache=TRUE}
m1 <- glmer(cbind(incidence, size-incidence) ~ period + (1|herd),
                  cbpp, binomial)
m1.z <- zeta(m1)
```

```{r fig2}
plot(m1.z)
```

```{r fig3}
plot(m1.z,scaled=TRUE)
```

Now we need a model with a vector-valued RE for an example.
I tried simulating something  before, but that appears (??)

```{r toenailfit,cache=TRUE}
toenail <- read.csv("toenail.csv")
m2 <- glmer(outcome~treatment+visit+(visit|patient),toenail,
            family=binomial)
```

```{r toenailzeta,cache=TRUE}
m2.z <- zeta(m2,grps=1:25)
```

This is a nice collection of mussels ...
```{r}
plot(m2.z)
```

Still trying to work out the analogue of Figure 3 (i.e.,
a version where we scale by the bivariate normal).
I think this actually *should* work, but this example
is very badly behaved in the corner ...

```{r ratios}
zz <- m2.z$zvals
mm <- 2*pi*dnorm2d(sqrt(m2.z$devarr[,,1]))
m0 <- 2*pi*dnorm2d(sqrt(outer(zz^2,zz^2,"+")))
par(mfrow=c(2,2))
persp(zz,zz,mm,col="gray",main="conditional density")
persp(zz,zz,m0,col="lightblue",main="bivariate normal")
persp(zz,zz,mm/m0,col="pink",main="ratio")
persp(zz,zz,log(mm/m0),col="lightgreen",main="log ratio")
```

```{r eval=FALSE}
## also wrong?
dnorm2dsq <- function(x) {
    dnorm(sqrt(x))/sqrt(2*pi)  ## == exp(-x^2/2)/(2*pi)
}

contourplot(dnorm2dsq(dev)/dnorm2dsq(sqrt(zval1^2+zval2^2)) ~ zval1*zval2|id,
             scales=list(z=list(relation="free")),
             data=m2.zm,
             labels=FALSE,
             type=c("g","l"), aspect=0.6, layout=c(5,5),
             xlab="z", ylab="density")
```

```{r echo=FALSE,eval=FALSE}
## old testing stuff
    ss <- subset(melt(m2.z),id=="1")
    ss0 <- m2.z$devarr[,,1]
    contour(ss0)
    contour(matrix(exp(-ss$dev/2)/(2*pi),nrow=31))
    contourplot(exp(-dev/2)/(2*pi)~zval1*zval2,data=ss)
```


