Refit weirdness
===============

Exploring/expanding [issue #316](https://github.com/lme4/lme4/issues/316) from @pitakakariki:

```{r pkgs}
library("lme4")
library("plyr")   ## for ldply
```

Simulation setup:
```{r sim1}
set.seed(123)
n <- c(2765,439,3356,82,801,798,329,299,319,2388,128,252,190,789,4322,92)
f <- as.factor(rep(letters[1:16], n))
x <- sample(1:7, sum(n), replace=TRUE)
obs <- 1:sum(n)
testdata <- data.frame(x, f, obs)
p <- list(theta=c(0.89, 0, 1.04), beta=c(-0.46, 0.0048))
testdata$y <- suppressWarnings(simulate( ~ x + (x||f) + (1|obs),
                                        family="poisson",
                       newdata=testdata, newparams=p)[[1]])
```

```{r fit1,cache=TRUE}
t1 <- system.time(glmer_orig <- glmer(y ~ x + (x||f) + (1|obs),
                                      family="poisson", data=testdata))
z <- simulate(glmer_orig,seed=42)[[1]]
```

Summary function: give log-likelihood and deviance relative to
first value, elapsed time, and estimated fixed-effect parameter values ...

```{r sumfun}
cfun <- function(x.list,t.list) {
    dfun <- if (getME(x.list[[1]],"REML")==0) deviance else REMLcrit
    tvec <- sapply(t.list,"[","elapsed")
    glance.list <- ldply(x.list,function(x) c(logLik=logLik(x),dev=dfun(x)))
    glance.list[,-1] <- lapply(glance.list[,-1],function(x) x-x[1])
    glance.list <- cbind(glance.list,t=tvec)
    comb.list <- merge(glance.list,ldply(x.list,fixef))
    return(comb.list)
}
```

## Fit `glmer()` models

```{r glmerfits,cache=TRUE}
t2 <- system.time(glmer_refit0 <- refit(glmer_orig, z))
t3 <- system.time(glmer_refit_start <- refit(glmer_orig,
                                             z, start=list(theta=c(1,1,1))))
t4 <- system.time(glmer_update <- update(glmer_orig,
                                         data=transform(testdata, y=z)))
```

```{r sumglmerfits}
glmer.list <- lme4:::namedList(glmer_refit0,
                               glmer_refit_start,glmer_update)
t.list <- list(t2,t3,t4)
options(digits=3)
cfun(glmer.list,t.list)
```

Uh-oh.

* `refit` with and without setting the starting values back to
the original values get (almost) the same answers, but using `update()`
instead gives a radically different result for the deviance
*despite the fact that the log-likelihood is off by only 0.14 ...*

## Fit `lmer()` models

```{r lmerfits,cache=TRUE}
system.time(lmer_orig <- lmer(y ~ x + (x||f) + (1|obs), data=testdata,
            control=lmerControl(check.nobs.vs.nlev="ignore",
                                check.nobs.vs.nRE="ignore"),
                              REML=TRUE))
t5 <- system.time(lmer_refit0 <- refit(lmer_orig, z))
t6 <- system.time(lmer_refit_start <- refit(lmer_orig,
                                            z, start=list(theta=c(1,1,1))))
t7 <- system.time(lmer_update <- update(lmer_orig,
                                        data=transform(testdata, y=z)))
```

```{r sumlmer}
lmer.list <- lme4:::namedList(lmer_refit0,lmer_refit_start,lmer_update)
t2.list <- list(t5,t6,t7)
cfun(lmer.list,t2.list)
```

Can we refactor `refit`?

```{r}
refit2 <- function(object,y) {
    newobj <- object
    ## make new resp mod
    ## how do I step back just far enough to do this?
    ## extract reTrms equivalent from fitted model
    ## modify model frame?
    mkLmerDevfun(fr, X, reTrms, REML = TRUE, start = NULL,
                 verbose = 0, control = lmerControl(), ...)

}
```


