> I guess lme4 might have some magic to temporarily suppress the warnings in summary() so withCallingHandlers() is unable to catch them. I looked at the source code of lme4 for a while, but did not figure out whether or how they did it. Here is a reproducible example that you should report to https://github.com/lme4/lme4/issues (ping @bbolker or @mmaechler or someone in the dev team):

A simpler example than the one presented:

```{r setup}
library("lme4")
ss <- transform(sleepstudy,obs=factor(1:nrow(sleepstudy)))
```

Regular run produces a warning:

```{r lme4_warn}
m1 <- lmer(Reaction~1+(1|obs),data=ss,
     control=lmerControl(check.nobs.vs.nlev="warning",
                         check.nobs.vs.nRE="ignore"))
```

Running inside `summary()` doesn't:

```{r lme4_warnsum}
sum1 <- summary(m1 <- lmer(Reaction~1+(1|obs),data=ss,
     control=lmerControl(check.nobs.vs.nlev="warning",
                         check.nobs.vs.nRE="ignore")))
```

**BUT** this is the same behaviour as base-R functions!

`sqrt()` produces a warning in this case:

```{r sqrt_warn}
sq <- sqrt(-1)
```

... but if we nest inside `summary()` it doesn't!

```{r sqrt_warnsum}
summary(sq <- sqrt(-1))
```
